Lexical Richness as a Lens: Exploring the Influence of Social Class on Authors' Writings

### 2024-03-17 paper/paper.qmd
ref, intro fixed

# Introduction {#sec-intro}
At the end of the 20th century, the quantitative study of vocabulary in texts reached a level of precision that spurred a series of linguistic studies. Vocabulary is a fundamental part of language and is the unit of meaning that makes up larger structures such as sentences, paragraphs, and full texts [@cite2020vocabulary]. The use of vocabulary represents the author's writing style and is also influenced by family education and social class. The question of whether the vocabulary used by different authors in their writing is influenced by their class of origin deserves to be explored in depth. The purpose of this paper is to gain insight into whether class differences affect the richness of vocabulary and central ideas in works by examining whether upper-class authors use richer writing vocabulary in their essays than middle-class authors.

Utilizing a combination of quantitative data analysis and authorship studies, we investigated the Type-Token Ratio of upper- and middle-class authors to reflect the richness of textual vocabulary over the period 1930-1990. This richness gives us an idea of the number of different terms used in a text and the diversity of the vocabulary [@cite2013lexical]. For the purpose of this study, we carefully selected two authors from the upper class, namely Nancy Mitford and Anthony Powell, alongside two authors from the working class, Hugh Garners and David Herbert Lawrence, as subjects of our investigation. In order to fulfill the objectives of our study, we employ the method of calculating the Corrected Type-Token Ratio (CTTR) for lexical analysis, as this approach offers a more accurate account of the texts produced by the four selected authors. The type token ratio is calculated for data files containing transcribed utterances. In language acquisition research it often means transcriptions of spontaneous speech, including unguided narratives, guided retellings, or speech elicited by a series of successive pictures [@cite2007comparing]. A higher Corrected Type-Token Ratio (CTTR) indicates that the author possesses a richer vocabulary, demonstrating a greater diversity in word usage within their texts. The CTTR values of four writers from distinct social classes were analyzed to assess the richness of their vocabulary usage. The comparison revealed that the works of upper-class writers exhibited higher CTTR values relative to those of middle-class writers, suggesting a more diverse vocabulary in the literary creations of the upper-class authors. This phenomenon suggests that different classes of writers have different writing styles and preferred vocabularies, and that the writers' writing styles and ideas in their works are influenced by their vocabulary choices.

@@

## Estimand
In this study, our main object of estimation is the causal effect of different classes on the vocabulary use of writers' writing, i.e., changes that can be attributed to the writers' own class of origin, divorced from the economic factors of their times. This involves analyzing the richness of the writing vocabulary of authors from both the upper and middle classes, in order to understand how differences in education, life, and personal perceptions brought about by different class origins affect writers' ideas and writing styles.

@@

Data on word counts and unique word counts for the four selected authors were sourced from the original text files of the selected authors' writings. The original files of these writings were gathered through online library sources, these files were processed and transformed into workable data using R[@citeR]. For key operations, please refer to the [@sec-appendix].

@@

#Model

@@

@sec-result

@@

@tbl-author-cttr-dist shows the average Corrected Type-Token Ratio (CTTR) and the associated standard deviation for each of the four studied authors. David Herbert Lawrence and Nancy Mitford both have high average CTTRs at 31.61 and 31.97 respectively, indicating a greater lexical richness in their works. However, Lawrence's works show more variability with a standard deviation of 7.37, as opposed to Mitford's 1.49, suggesting her use of vocabulary is more consistent. Hugh Garners has a notably lower average CTTR of 16.31, paired with a standard deviation of 2.31, which is indicative of a consistently narrower range of vocabulary. Powell Anthony tops the average CTTR at 34.67 with a standard deviation of 1.64, demonstrating not only the highest lexical richness but also a high consistency in vocabulary usage across his works. These figures collectively offer a quantitative insight into the diversity and consistency of language used by authors from different social backgrounds.

@@

@tbl-multi-lev-model provides coefficients from a negative binomial regression and a multilevel negative binomial regression, indicating how each author's work differs from the intercept in terms of Corrected Type-Token Ratio (CTTR). Hugh Garners shows a negative coefficient (-0.658 in Neg Binom and -1.308 in Multilevel Neg Binom), suggesting his CTTR is significantly lower than the baseline set by the intercept, which is consistent with the shortest average word count observed in the previous analysis. Nancy Mitford's coefficient is slightly above zero (0.021 in Neg Binom and 0.152 in Multilevel Neg Binom), indicating her CTTR is almost at the baseline level. This aligns with her position as neither the lowest nor the highest in terms of lexical richness. Powell Anthony’s positive coefficient (0.107 in Neg Binom and 0.153 in Multilevel Neg Binom) implies his CTTR is above the baseline, though not as high as David Herbert Lawrence's. David Herbert Lawrence does not have a coefficient listed here but is included in the distribution figure below.

@@

To gain a more intuitive understanding of the model, we visualized the coefficients from @tbl-multi-lev-model in the form of a distribution figure. @fig-multi-lev-model-by-author, presents the distribution of CTTR estimates for each author, providing a visual representation of the variance around the mean estimate. Powell Anthony’s distribution is relatively narrow and centered around the mean, suggesting his lexical richness is consistently above the baseline with less variation in CTTR scores. Nancy Mitford's distribution also clusters closely around the mean but shows a slight skew towards higher values in the multilevel negative binomial regression, indicating a modest variability with a tendency for higher CTTR scores. Hugh Garners' distribution is broader and notably shifted to the left of the baseline in both models, implying a lower CTTR and greater variability in lexical richness compared to the others. David Herbert Lawrence, while not explicitly compared against a coefficient in the table, is represented in the figure with the broadest distribution. This suggests that his CTTR is highly variable, confirming the previous analysis which indicated his works are both lengthy and diverse in terms of lexical richness.
Overall, the @fig-multi-lev-model-by-author corroborates the numerical findings of @tbl-multi-lev-model, offering a visual confirmation of the differences in lexical richness and variability among the authors. It highlights the significant spread in the lexical diversity of Lawrence's works, as well as the relative consistency observed in the writings of Powell Anthony and Nancy Mitford, with Garners' works showing both a lower mean and a higher dispersion.

@@

David Herbert Lawrence is a conspicuous outlier in this analysis. His works are scattered across a wide range of both CTTR and word counts, showing that while he has a propensity for longer works, the lexical richness within those works is quite variable. Hugh Garners’ points cluster at the lower end of both word count and CTTR scales, suggesting that his works are shorter and less lexically rich. This could reflect the impact of social and educational limitations on his writing style. Nancy Mitford's works are positioned in the middle range of the word counts but show a higher CTTR, similar to that of Lawrence, indicating a richer vocabulary that is not necessarily dependent on the length of the text. Powell Anthony stands out with a group of works that not only have high CTTR values but also fall into a moderate range of word counts. His data points suggest a strong consistency in lexical richness that does not fluctuate widely with the length of the text.

@@

## Social and historical influences

@@

# Data Manipulation and Cleaning {#sec-appendix}

In the context of our raw data, we are dealing with raw text files of authors' writings. Thus, we begin by omitting unnecessary portions (such as publishers' notes) the text files. To transform these raw texts to workable data, we created three functions, `word_dist()`, `cttr_dist()`, and `cttr()` using R libraries [@citetidyverse, @citedplyr, @citehere, @citekableExtra]. These functions aim to gather the descriptive statistics of each writings partially (average, mean, cttr), and store them into `data/analysis_data` with the author name.

@@

@tbl-multi-lev-model and @fig-multi-lev-model-by-author were created using the imports from `models/author_cttr_model_multi.rds` and `models/author_cttr_model.rds`. These `.rds` files were created in `scripts/03-models.R` using functions such as `stan_glm()` from R libraries [@citetidyverse, @citerstanarm]. Moreover, the `corrected_type_token_ratio` from `data/analysis_data/cttr_all.csv` were rounded to integers in order to use Negative Binomial regression.

@@

>>>>

# Introduction {#sec-intro}
By the late 20th century, the quantitative study of vocabulary in texts reached a level of precision that spurred a series of linguistic studies. Vocabulary is a fundamental part of language and is the unit of meaning that makes up larger structures such as sentences, paragraphs, and full texts [@cite2020vocabulary]. The use of vocabulary represents the author's writing style and is also influenced by family education and social class. The purpose of this paper is to gain insight into whether class differences affect the richness of vocabulary and central ideas in works by examining whether upper-class authors use richer writing vocabulary in their essays than middle-class authors.

Utilizing a combination of quantitative data analysis and authorship studies, we investigated the Type-Token Ratio of upper- and middle-class authors to reflect the richness of textual vocabulary over the period 1930-1990. This richness gives us an idea of the number of different terms used in a text and the diversity of the vocabulary[@cite2013lexical]. For this study, we selected two upper-class authors, Nancy Mitford and Anthony Powell, and two working-class authors, Hugh Garners and David Herbert Lawrence. We utilized the Corrected Type-Token Ratio (CTTR) for a detailed lexical analysis of their texts. In language acquisition studies, this often entails transcriptions of spontaneous speech, such as narratives, retellings, and picture-elicited speech[@cite2007comparing]. A higher Corrected Type-Token Ratio (CTTR) indicates that the author possesses a richer vocabulary, demonstrating a greater diversity in word usage within their texts. This phenomenon suggests that different classes of writers have different writing styles and preferred vocabularies, and that the writers' writing styles and ideas in their works are influenced by their vocabulary choices.

@@

## Estimand
This study focuses on estimating the causal effect of class origin on writers' vocabulary use, specifically how it shapes and changes in writing style. By examining the vocabulary richness of authors from upper and working classes, we aim to understand how variations in education, life experiences, and personal perceptions due to class differences influence writing.

@@

Data on word counts and unique word counts for the four selected authors were sourced from the original text files of the selected authors' writings. The original files of these writings were gathered through online library sources (please refer to [@sec-appendixB]), these files were processed and transformed into workable data using R[@citeR]. For key operations, please refer to the [@sec-appendixA].

@@

# Model {#sec-model}

@@

@sec-result presents the core findings on the relationship between authors' word count and CTTR, specifically focusing on the influence of social class and educational background on literary vocabulary richness.

@@

@tbl-author-cttr-dist shows the average Corrected Type-Token Ratio (CTTR) and the associated standard deviation for each of the four studied authors.

David Herbert Lawrence and Nancy Mitford both have high average CTTRs at 31.61 and 31.97 respectively, indicating a greater lexical richness in their works. However, Lawrence's works show more variability with a standard deviation of 7.37, as opposed to Mitford's 1.49, suggesting her use of vocabulary is more consistent.

Hugh Garners has a notably lower average CTTR of 16.31, paired with a standard deviation of 2.31, which is indicative of a consistently narrower range of vocabulary. Powell Anthony tops the average CTTR at 34.67 with a standard deviation of 1.64, demonstrating not only the highest lexical richness but also a high consistency in vocabulary usage across his works.

These figures collectively offer a quantitative insight into the diversity and consistency of language used by authors from different social backgrounds.

@@

@tbl-multi-lev-model provides coefficients from a negative binomial regression and a multilevel negative binomial regression, indicating how each author's work differs from the intercept in terms of Corrected Type-Token Ratio (CTTR).

Hugh Garners shows a negative coefficient (-0.658 in Neg Binom and -1.308 in Multilevel Neg Binom), suggesting his CTTR is significantly lower than the baseline set by the intercept, which is consistent with the shortest average word count observed in the previous analysis.

Nancy Mitford's coefficient is slightly above zero (0.021 in Neg Binom and 0.152 in Multilevel Neg Binom), indicating her CTTR is almost at the baseline level. This aligns with her position as neither the lowest nor the highest in terms of lexical richness.

Powell Anthony’s positive coefficient (0.107 in Neg Binom and 0.153 in Multilevel Neg Binom) implies his CTTR is above the baseline, though not as high as David Herbert Lawrence's.

David Herbert Lawrence does not have a coefficient listed here but is included in the distribution figure below.

@@

To gain a more intuitive understanding of the model, we visualized the coefficients from @tbl-multi-lev-model in the form of a distribution figure. @fig-multi-lev-model-by-author, presents the distribution of CTTR estimates for each author, providing a visual representation of the variance around the mean estimate.

Powell Anthony’s distribution is relatively narrow and centered around the mean, suggesting his lexical richness is consistently above the baseline with less variation in CTTR scores. Nancy Mitford's distribution also clusters closely around the mean but shows a slight skew towards higher values in the multilevel negative binomial regression, indicating a modest variability with a tendency for higher CTTR scores.

Hugh Garners' distribution is broader and notably shifted to the left of the baseline in both models, implying a lower CTTR and greater variability in lexical richness compared to the others. David Herbert Lawrence, while not explicitly compared against a coefficient in the table, is represented in the figure with the broadest distribution. This suggests that his CTTR is highly variable, confirming the previous analysis which indicated his works are both lengthy and diverse in terms of lexical richness.

Overall, the @fig-multi-lev-model-by-author corroborates the numerical findings of @tbl-multi-lev-model, offering a visual confirmation of the differences in lexical richness and variability among the authors. It highlights the significant spread in the lexical diversity of Lawrence's works, as well as the relative consistency observed in the writings of Powell Anthony and Nancy Mitford, with Garners' works showing both a lower mean and a higher dispersion.

@@

David Herbert Lawrence is a conspicuous outlier in this analysis. His works are scattered across a wide range of both CTTR and word counts, showing that while he has a propensity for longer works, the lexical richness within those works is quite variable. Hugh Garners’ points cluster at the lower end of both word count and CTTR scales, suggesting that his works are shorter and less lexically rich. This could reflect the impact of social and educational limitations on his writing style.

Nancy Mitford's works are positioned in the middle range of the word counts but show a higher CTTR, similar to that of Lawrence, indicating a richer vocabulary that is not necessarily dependent on the length of the text. Powell Anthony stands out with a group of works that not only have high CTTR values but also fall into a moderate range of word counts. His data points suggest a strong consistency in lexical richness that does not fluctuate widely with the length of the text.

@@

## Social and Historical Influences

@@

# Data Manipulation and Cleaning {#sec-appendixA}

In the context of our raw data, we are dealing with raw text files of authors' writings. Thus, we begin by omitting unnecessary portions (such as publishers' notes) the text files. To transform these raw texts to workable data, we created three functions, `word_dist()`, `cttr_dist()`, and `cttr()` using R libraries [@citetidyverse; @citedplyr; @citehere; @citekableExtra]. These functions aim to gather the descriptive statistics of each writings partially (average, mean, cttr), and store them into `data/analysis_data` with the author name.

@@

@tbl-multi-lev-model and @fig-multi-lev-model-by-author were created using the imports from `models/author_cttr_model_multi.rds` and `models/author_cttr_model.rds`. These `.rds` files were created in `scripts/03-models.R` using functions such as `stan_glm()` from R libraries [@citetidyverse; @citerstanarm]. Moreover, the `corrected_type_token_ratio` from `data/analysis_data/cttr_all.csv` were rounded to integers in order to use Negative Binomial regression.

@@

# Referenced Writings {#sec-appendixB}

[@mitford2011penguin; @mitford2015don; @mitford2001madame; @mitford1970frederick; @mitford2010love; @mitford2012sun; @mitford2010pursuit; @powell1977kindly; @powell1988buyer; @powell1968military; @powell2011temporary; @powell2010hearing; @powell1994valley; @powell1957lady; @powell1955acceptance; @powell2011books; @powell1966soldier; @powell2011casanova; @lawrence1992virgin; @lawrence2005rainbow; @lawrence2022plumed; @lawrence2012sons; @lawrence2013odour; @lawrence2007lady; @lawrence2002kangaroo; @lawrence2015england; @garner2003hugh; @garner2014waste]
